A Pitman-Yor process prior for estimating shared evolutionary events

### 2022-06-06 manuscript/ms.tex
Update on Overleaf.

\begin{abstract}
Evolutionary biologists are interested in investigating the evolutionary history of the species. Models are one of the means that allow them to model the world or the evolutionary processes. Therefore, knowing the behavior of the models is important. This study compares the performance of three models Dirichlet process, Pitman-Yor process, and weighted-uniform(DP, PYP, and WU) that are implemented in \ecoevolity software to estimate the number of shared divergence time among 10 pairs of population in the past. Our results show that Pitman-Yor process which is the more flexible model, consistently outperforms the other two models in both simulations (with reduced data sets). The flexibility of Pitman-Yor process has some advantages, especially when it is wrong.

@@

\section{Introduction}
Rewrite Intro because it's kind of overlapping with the result section.

Phylogeographic Study \\
Sharing divergence time across taxa \\
Shared divergence across taxa happened via geological events. \\
Temporal clustering reveals evolutionary history of the taxa. \\
Temporal pattern- \\
Model the evolutionary process \\
Important to know the behavior of the models \\
Evolutionary events

We compared the performance of 3 models (Dirichlet prior, Pitman-Yor prior and Uniform prior) 
that are implemented in Ecoevolity software which is the primary software we used to estimate the number of shared divergence time 
among 10 pairs of populations in the past. We simulated data and know the true number of divergence time from it. We are interested 
to estimate the number of divergence time from the generated data. We generated data under 5 various ways to putting prior over all
of the ways that we can assign a pair to divergence event. And then we analyzed the data under 3 models that we can model  
and have some sense of comparison. We are interested in how well the models have estimated the divergence time and how well 
the models have estimated what population pairs belong to which divergence time. We assigned prior probabilities to 3 models; how probable each model is. Dirichlet-process prior 
assigns prior to a new category with probability of \(\dfrac{1}{\concentration +1}\) and an existing category with probability of \(\dfrac{\concentration}{\concentration+1}\). The concentration parameter(\concentration) 
makes the DP flexible because we can adjust the concentration parameter to fit our prior expectation. We can put a distribution on the 
concentration parameter and integrate over uncertainty about the prior probabilities of the divergence models. Pitman-Yor process prior is 
generalization of Dirichlet process. There is a discount parameter($\discount$) added in PYP and when $\discount= 0$ PYP converges to DP. Dirichlet has an exponential tail 
and the discount parameter gives the PYP flexibility over the tail behavior. Uniform prior is every possible divergence model is equally probable. 
Then we have specified the priors in the config files for each of these 3 models that they are comparable to each other to assess the behavior of the models.

We used these 3 config files and then added 2 more configs (independent and simutanous configs) to simulate data sets. Independent config 
basically specifies 10 divergence times and each population diverged independently. Simutanous config specifies that the populations 
diverged simutanously and there is 1 shared event. And there is another config file which ignores the constant sites (characters). A total of 6 config 
files are specified to simulate the data sets. We used these 6 config files 4 times to generate scripts and then simulated 10 data sets under each of 
these 5 models with 5000,000 characters. Then we analyzed the data under 3 models and the results show that Pitman-Yor process is doing better job handling 
most difficult scenario all 10 pairs diverged independently because PYP is more flexible model. The coverage is more robust for Pitman-Yor process twice likely.
So we are interested to see whether this pattern holds with less data. We decided to do another simulation with 4 varying data sets 5,000, 10,000, 50,000, and 100,000. 
All the data sets simulated under 5 models and then analyzed these data sets under 3 models (DP, PYP, and SW). There are equal number of replicates for both simulations 
a total of 720 replicates for each simulation. We found that with varying data sets suggest the same Pitman-Yor process has 2 parameters and more flexible might 
have some advantages especially when it is wrong.

@@

We used \simcoevolity and \ecoevolity of the \ecoevolity software
package
\citep[Version 0.3.2 Commit c1685dfa][]{Oaks2018ecoevolity}
to simulate and analyze \datasets, respectively.
Each simulated \dataset comprised biallelic characters from 2 diploid
individuals (4 genomes) sampled per population from 10 pairs of populations
(i.e., 10 divergence comparisons).
We assumed the mutation rates were equal and 1 across the 10 pairs of
populations, such that time and effective population sizes are scaled by the
mutation rate.
The times of divergence events were exponentially distributed with a mean 0.05
expected subsitutions per site.
The mutation-scaled effective sizes ($\epopsize\murate$) of the descendant
populations were gamma-distributed with a shape of 20 and mean of 0.001,
and the size of the ancestral population relative to the mean of its
descendants was gamma distributed with a shape of 50 and mean of 1.
These distributions were also used as priors when the simulated \datasets were
analyzed with \ecoevolity.

@@

to look at the results of divegence time estimation with constant sites. The models which generated 
the data in rows and columns represent the models under which the data is analyzed. The true divergene 

@@

time versus estimated divengence time show that all the three models (DP, PYP, and SW) performed
equally well with very similiar coverage around 95\%. Except for independent all others have truths 
contianed in within that 95\% credible interval. Independent is little lower because it's evolving 
indepdently and we are not modeling that way. We haven't observed any patterns that differiate
these 3 models. Root mean square errors are very similiar across all the models as well. 

@@

sites which contain lot of information and when we use them we have much more presicion in estimates
than ingoring them. All 10 pairs diverge spontatnous recently causing some MCMC issues and there 
are very few variable sites all 10 pairs diverged recently. Basicailly there is discrepency bewtween 

@@

likelihood for the missing invariable characters is corrected in \Ecoevolity. PYP is doing better when 

>>>>

\begin{abstract}
Evolutionary biologists are interested in investigating the evolutionary history of the species. phylogenetic models are one of the means that allow them to model the world or the evolutionary processes that are responsible for generating biodiversity patterns across the tree of life. These models also be able to predict or test the primary mechanism for population divergence. Therefore, knowing the behavior of the models is important. This study compares the performance of three models Dirichlet process, Pitman-Yor process, and weighted-uniform(DP, PYP, and WU) that are implemented in \ecoevolity software to estimate the number of shared divergence time among 10 pairs of population in the past.Our results show that Pitman-Yor process which is the more flexible model, consistently outperforms the other two models in both simulations (with reduced data sets). The flexibility of Pitman-Yor process has some advantages, especially when it is wrong.

@@

\section{Introduction}
The primary goal of biogeography is to understand how the large-scale evolutionary processes like climate change affect the diversification. Evolutionary processes have a large effect on the diversity and diversification of the species. Biogeographers are interested in knowing the biogeographic patterns among the co-distributed species. How these large-scale evolutionary processes affect the species' evolutionary history and what kind of roles these evolutionary processes play in creating the biodiversity patterns across the taxa. Climate change is one of processes that shapes the evolutionary history of the species. Climate fluctuated through the earth's history from warming to cold, glaciation to deglaciation, and arise in sea level vice verse. For some species, climate-driven mechanism is the primary mode of population divergence \citep{Oaks2019codemogpreprint}.\\ Phylogenetic models allow to capture micro and macroevolutionary processes that have generated biodiversity patterns across the tree of life \citep{Oaks2022}. Various phylogenetic models from generalized tree model to process-based models can be utilized to predict or test for the underlying mechanisms which created a diverse life on the earth. Estimating shared evolutionary events is an interest of many sub-fields of biology (biogeography etc.,). Biogeographers are interested in addressing questions like the number of shared divergence times, the timing of divergence, and whether there is a concordance between geological and biological events across the taxa. The usage of phylogenetic models in various fields of biology is wide. Therefore, knowing the behavior of models is vital for studies which depend on models to test hypotheses and obtain the results.\\
In this study, we compared the performance of 3 models (Dirichlet prior, Pitman-Yor prior and Uniform prior) 
that are implemented in \ecoevolity software which is the primary software we used to estimate the number of shared divergence time among 10 pairs of populations in the past. We are interested in how well the models have estimated the shared divergence times and how well the models have estimated which population pairs belong to which divergence time. Inferring the shared divergence time across pairs of taxa can reveal the evolutionary history of taxa. It also shows what shared mechanism is the driving force for population divergence. Temporal patterns

We assigned prior probabilities to 3 models; how probable each model is. Dirichlet-process prior 
assigns prior to a new category with probability of \(\dfrac{1}{\concentration +1}\) and an existing category with probability of \(\dfrac{\concentration}{\concentration+1}\). The concentration parameter(\concentration) 
makes the DP flexible because we can adjust the concentration parameter to fit our prior expectation. We can put a distribution on the 
concentration parameter and integrate over uncertainty about the prior probabilities of the divergence models. Pitman-Yor process prior is 
generalization of Dirichlet process. There is a discount parameter($\discount$) added in PYP and when $\discount= 0$ PYP converges to DP. Dirichlet has an exponential tail 
and the discount parameter gives the PYP flexibility over the tail behavior. Uniform prior is every possible divergence model is equally probable. 
Then we have specified the priors in the config files for each of these 3 models that they are comparable to each other to assess the behavior of the models.

We used these 3 config files and then added 2 more configs (independent and simultaneous configs) to simulate data sets. Independent config 
basically specifies 10 divergence times and each population diverged independently. Simultaneous config specifies that the populations 
diverged simultaneously and there is 1 shared event. And there is another config file which ignores the constant sites (characters). A total of 6 config 
files are specified to simulate the data sets. We used these 6 config files 4 times to generate scripts and then simulated 10 data sets under each of 
these 5 models with 5000,000 characters. Then we analyzed the data under 3 models and the results show that Pitman-Yor process is doing better job handling 
most difficult scenario all 10 pairs diverged independently because PYP is more flexible model. The coverage is more robust for Pitman-Yor process twice likely.
So we are interested to see whether this pattern holds with less data. We decided to do another simulation with 4 varying data sets 5,000, 10,000, 50,000, and 100,000. 
All the data sets simulated under 5 models and then analyzed these data sets under 3 models (DP, PYP, and SW). There are equal number of replicates for both simulations 
a total of 720 replicates for each simulation. We found that with varying data sets suggest the same Pitman-Yor process has 2 parameters and more flexible might 
have some advantages especially when it is wrong.

@@

We used \simcoevolity and \ecoevolity of the \ecoevolity software
package
\citep[Version 0.3.2 Commit c1685dfa][]{Oaks2018ecoevolity}
to simulate and analyze \datasets, respectively.
Each simulated \dataset comprised biallelic characters from 2 diploid
individuals (4 genomes) sampled per population from 10 pairs of populations
(i.e., 10 divergence comparisons).
We assumed the mutation rates were equal and 1 across the 10 pairs of
populations, such that time and effective population sizes are scaled by the
mutation rate.
The times of divergence events were exponentially distributed with a mean 0.05
expected substitutions per site.
The mutation-scaled effective sizes ($\epopsize\murate$) of the descendant
populations were gamma-distributed with a shape of 20 and mean of 0.001,
and the size of the ancestral population relative to the mean of its
descendants was gamma distributed with a shape of 50 and mean of 1.
These distributions were also used as priors when the simulated \datasets were
analyzed with \ecoevolity.

@@

to look at the results of divergence time estimation with constant sites. The models which generated 
the data in rows and columns represent the models under which the data is analyzed. The true divergence 

@@

time versus estimated divergence time show that all the three models (DP, PYP, and SW) performed
equally well with very similar coverage around 95\%. Except for independent all others have truths 
contained in within that 95\% credible interval. Independent is little lower because it's evolving 
independently and we are not modeling that way. We haven't observed any patterns that differentiate
these 3 models. Root mean square errors are very similar across all the models as well. 

@@

sites which contain lot of information and when we use them we have much more precision in estimates
than ignoring them. All 10 pairs diverge spontaneous recently causing some MCMC issues and there 
are very few variable sites all 10 pairs diverged recently. Basically there is discrepancy between 

@@

likelihood for the missing invariable characters is corrected in \ecoevolity. PYP is doing better when 
